LPME Simulation Studies

### 2025-02-10 code/01_lpme_simulations.qmd
align with reproducibility template

---
title: "LPME Simulation Studies"
author: "Robert Zielinski"
date: "April 23, 2024"
output: html_document
---

This notebook includes all code necessary to replicate the results of the simulations comparing performance between the LPME, PME, and PC/PS algorithms on data generated from a common underlying manifold, but with varying degrees of noise added between time points. The LPME algorithm is designed to obtain *smooth* estimates of the manifolds underlying noisy data. Consequently, if the algorithm achieves its stated goal, it will not show the best goodness of fit to the observed data in these conditions. Simulated data is useful in this setting because we can generate the data from a known underlying manifold, allowing us to measure the algorithm's success at recovering this known manifold.

```{r, message=FALSE}
library(tidyverse)
library(plotly)
library(pracma)
library(foreach)
library(doParallel)
library(furrr)
library(progress)
library(pme)
library(princurve)

# note that file paths are relative to the data/ directory where this notebook resides
source("functions/sim_data.R")
source("functions/calc_pme_est.R")
source("functions/calc_lpme_est.R")
source("prinSurf_v3.R")
```

Manifold learning algorithms traditionally have tradeoffs when compared to each other, meaning that an algorithm that displays strong performance for one manifold may be outperformed by another algorithm when considering a different manifold. Because of this, manifold learning algorithms are often evaluated with several manifolds meant to test an algorithm's performance on data with various attributes. We consider eight different manifolds:

1. For $r \in [-3, 3]$, $f(r) = (r, \sin(r + \frac{\pi}{2}))$.

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 1,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 1],
  z = sim_df[, 3],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 1],
    z = true_vals[, 3],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 1")
```

2. For $r \in [-3\pi, 3\pi]$, $f(r) = (r, \sin(r))$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 2,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 1],
  z = sim_df[, 3],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 1],
    z = true_vals[, 3],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 2")
```

3. For $r \in [-\frac{4}{5}\pi, \frac{\pi}{2}]$, $f(r) = (\cos(r), \sin(r))$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 3,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 1],
  z = sim_df[, 3],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 1],
    z = true_vals[, 3],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 3")
```

4. For $r \in [-1, 1]$, $f(r) = (r, r^2, r^3)$

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 4,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 4")
```

@@

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 5,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 5")
```

@@

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 6,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 6")
```

@@

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 7,
  noise = 0.15,
  amp_noise = 0.25 / 25, # noise reduced to maintain integrity of shape
  period_noise = 0.25 / 25,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 7")
```

@@

```{r, echo=FALSE}
set.seed(500)
time_vals <- seq(0, 1, 0.2)
sim_list <- lapply(
  time_vals,
  sim_data,
  case = 8,
  noise = 0.15,
  amp_noise = 0.25,
  period_noise = 0.25 / 5,
  N = 1000,
  time_change = 0,
  time_trend = "constant"
)

sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
for (i in 1:length(sim_list)) {
  sim_df <- rbind(sim_df, sim_list[[i]][[1]])
  true_vals <- rbind(true_vals, sim_list[[i]][[2]])
}
sim_df <- sim_df[-1, ]
true_vals <- true_vals[-1, ]

plot_ly(
  x = sim_df[, 2],
  y = sim_df[, 3],
  z = sim_df[, 4],
  frame = sim_df[, 1],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) %>%
  add_markers(
    x = true_vals[, 2],
    y = true_vals[, 3],
    z = true_vals[, 4],
    frame = true_vals[, 1],
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  ) %>%
  layout(title = "Simulation Case 8")
```

Simulation runs were conducted using functions for each case, with each function taking the following arguments:

- `max_time`: The maximum time from baseline of a set of observations, or the study duration.
- `interval`: The interval of study visits. Intervals between study visits were assumed to be constant in these simulations, but the LPME algorithm can handle settings with varying intervals between visits.
- `amp_noise`: Specifies the magnitude of between-time point noise impacting the amplitude of the functions used to generate the data.
- `shape_noise`: Specifies the magnitude of between-time point noise impacting the shape, or period, of the functions used to generate the data.
- `time_change`: Specifies the magnitude of structural change in the underlying manifold over time.
- `time_trend`: Describes the type of longitudinal change in the underlying manifold. May take one of four values - constant, linear, quadratic, or sinusoidal.
- `n`: The number of observations generated at each time point.
- `initialization_alg`: The manifold learning algorithm used to initialize the PME algorithm. May currently take one of three values - isomap, diffusion_maps, or laplacian_eigenmaps.
- `run`: Used to save multiple simulation results with the same parameter inputs.
- `print_plots`: Specify whether plots of preliminary results should be viewed. This is useful for debugging but is computationally expensive.

The functions are given in the code chunks below:

### Simulation Case 1

```{r}
sim_error_case1 <- function(max_time, interval, amp_noise, shape_noise, time_change, time_trend, n, initialization_alg, run = 1, print_plots = FALSE) {
  time_vals <- seq(0, max_time, interval)
  # at each time point, simulate a dataset
  sim_list <- lapply(
    time_vals,
    sim_data,
    case = 1,
    noise = 0.15,
    amp_noise = amp_noise,
    period_noise = shape_noise,
    N = n,
    time_change = time_change,
    time_trend = time_trend
  )

  # combine the simulated observations into one matrix
  # do the same with points from the true manifold
  # these true points correspond to the simulated observations without added noise
  sim_df <- matrix(ncol = ncol(sim_list[[1]][[1]]))
  true_vals <- matrix(ncol = ncol(sim_list[[1]][[2]]))
  for (i in 1:length(sim_list)) {
    sim_df <- rbind(sim_df, sim_list[[i]][[1]])
    true_vals <- rbind(true_vals, sim_list[[i]][[2]])
  }
  sim_df <- sim_df[-1, ]
  true_vals <- true_vals[-1, ]

  # standardize observations to fall between -1 and 1 in each dimension
  for (col_idx in 1:ncol(sim_df)) {
    col_max <- max(abs(sim_df[, col_idx]))
    sim_df[, col_idx] <- sim_df[, col_idx] / col_max
    true_vals[, col_idx] <- true_vals[, col_idx] / col_max
  }

  # update time_vals to use standardized values
  time_vals <- unique(sim_df[, 1])

  # run lpme algorithm on observations
  lpme_result <- lpme(
    sim_df,
    1,
    smoothing_method = "spline",
    print_plots = print_plots,
    verbose = TRUE,
    init = "first",
    initialization_algorithm = initialization_alg
  )
  # compute values of observations projected to lpme estimated manifold
  lpme_vals <- calc_lpme_est(lpme_result, sim_df)

  # fit pme and principal curve estimates independently for each time point
  pme_result <- list()
  pme_vals <- list()
  # test all smoothing options for principal curve algorithm
  smoothing_options <- c("smooth_spline", "lowess", "periodic_lowess")
  principal_curve_result <- list()
  principal_curve_vals <- list()
  for (t in 1:length(time_vals)) {
    temp_data <- sim_df[sim_df[, 1] == time_vals[t], -1]
    pme_result[[t]] <- pme(
      temp_data,
      d = 1,
      initialization_algorithm = initialization_alg,
      initialization_type = "centers",
      verbose = FALSE
    )
    # save projections of observations onto PME manifold
    pme_vals[[t]] <- cbind(time_vals[t], calc_pme_est(pme_result[[t]], temp_data))
    curves <- list()
    curve_vals <- list()
    curve_error <- vector()
    for (smoother_idx in 1:length(smoothing_options)) {
      curves[[smoother_idx]] <- principal_curve(
        temp_data,
        smoother = smoothing_options[smoother_idx]
      )
      # save projections of observations onto estimated principal curve
      curve_vals[[smoother_idx]] <- cbind(time_vals[t], curves[[smoother_idx]]$s)
      curve_error[smoother_idx] <- curves[[smoother_idx]]$dist
    }
    opt_curve <- which.min(curve_error)
    # save principal curve results for smoothing approach with the lowest error
    principal_curve_result[[t]] <- curves[[opt_curve]]
    principal_curve_vals[[t]] <- curve_vals[[opt_curve]]
  }
  pme_vals <- reduce(pme_vals, rbind)
  principal_curve_vals <- reduce(principal_curve_vals, rbind)

  # create 3-dimensional plot of data, estimated manifolds, and true manifold
  p <- plot_ly(
    x = lpme_vals[, 2],
    y = lpme_vals[, 3],
    z = lpme_vals[, 1],
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3),
    name = "LPME"
  ) %>%
    add_markers(
      x = pme_vals[, 2],
      y = pme_vals[, 3],
      z = pme_vals[, 1],
      name = "PME"
    ) %>%
    add_markers(
      x = principal_curve_vals[, 2],
      y = principal_curve_vals[, 3],
      z = principal_curve_vals[, 1],
      name = "Principal Curve"
    ) %>%
    add_markers(
      x = true_vals[, 2],
      y = true_vals[, 3],
      z = true_vals[, 1],
      name = "True Manifold"
    ) %>%
    add_markers(
      x = sim_df[, 2],
      y = sim_df[, 3],
      z = sim_df[, 1],
      name = "Data",
      opacity = 0.2
    )

  # for each estimate, compute mean squared distance between
  # points on the true manifold and the same points projected to
  # the relevant manifold estimate
  pme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], pme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  principal_curve_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], principal_curve_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  lpme_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], lpme_vals[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

  data_error <- map(
    1:nrow(true_vals),
    ~ dist_euclidean(true_vals[.x, ], sim_df[.x, ])^2
  ) %>%
    unlist() %>%
    mean()

>>>>

---
title: "LPME Simulation Studies"
author: "Robert Zielinski"
date: "January 31, 2025"
format: html
---

This notebook includes all code needed to replicate the results of the simulations comparing performance between the Principal Manifold Estimation (PME), Longitudinal Principal Manifold Estimation (LPME), and Principal Curve/Surface (PC/PS) algorithms on data generated from a common underlying manifold, but with varying degrees of noise added independently between time points. The LPME algorithm is designed to obtain *smooth* estimates of the manifolds underlying noisy data. Consequently, if the algorithm achieves its stated goal, it will not show the best goodness of fit to the observed data in these conditions. Simulated data is useful in this setting because we can generate the data from a known underlying manifold, allowing us to measure the algorithm's success at recovering this known manifold.

```{r, echo=FALSE}
#| label: chunk_setup
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
#| label: load_packages
library(tidyverse)
library(plotly)
library(pracma)
library(foreach)
library(doParallel)
library(furrr)
library(progress)
library(pme)
library(princurve)
library(Rfast)

source("prinSurf_v3.R")
source("functions/simulations/simulate_data.R")
source("functions/simulations/preprocess_data.R")
source("functions/simulations/fit_models.R")
source("functions/simulations/evaluate_models.R")
source("functions/simulations/display_results.R")
source("functions/calculate_lpme_reconstructions.R")
source("functions/calculate_pme_reconstructions.R")
source("functions/estimate_volume.R")
source("functions/interior_identification.R")
```
## Simulated Manifolds

Machine learning algorithms traditionally have tradeoffs when compared to each other, meaning that an algorithm that displays strong performance for one manifold may be outperformed by another algorithm when considering a different manifold. Because of this, it is advantageous to evaluate manifold learning algorithms with several manifolds meant to test an algorithm's performance on data with a variety of attributes. The function given in the code chunk below will be used to generate data lying along the manifolds of interest.
In the function above, we consider eight different manifolds:

1. For $r \in [-3, 3]$, $f(r) = (r, \sin(r + \frac{\pi}{2}))$.
```{r, echo=FALSE}
#| label: plot_case1
set.seed(100)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 1,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~time,
  z = ~X2,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~time,
    z = ~X_true2,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

2. For $r \in [-3\pi, 3\pi]$, $f(r) = (r, \sin(r))$.
```{r, echo=FALSE}
#| label: plot_case2
set.seed(200)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 2,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~time,
  z = ~X2,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~time,
    z = ~X_true2,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

3. For $r \in [-\frac{4}{5}\pi, \frac{\pi}{2}]$, $f(r) = (\cos(r), \sin(r))$.
```{r, echo=FALSE}
#| label: plot_case3
set.seed(300)
sim_df <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 3,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~time,
  z = ~X2,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~time,
    z = ~X_true2,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

4. For $r \in [-1, 1]$, $f(r) = (r, r^2, r^3)$
```{r, echo=FALSE}
#| label: plot_case4
set.seed(400)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 4,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

@@

```{r, echo=FALSE}
#| label: plot_case5
set.seed(500)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 5,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )

```

@@

```{r, echo=FALSE}
#| label: plot_case6
set.seed(600)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 6,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )

```

@@

```{r, echo=FALSE}
#| label: plot_case7
set.seed(700)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 7,
  obs_noise = 0.15,
  amplitude_noise = 0.05,
  period_noise = 0.01,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )

```

@@

```{r, echo=FALSE}
#| label: plot_case8
set.seed(800)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 8,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )

```

Note that case 8, in which observations drawn from a 2-dimensional manifold take the form of a sphere in 3 dimensions, is of particular interest due to our application to the surfaces of subcortical structures in the brain.

Simulation runs were conducted by varying the following arguments:

- `duration`: The simulated study's follow-up duration. Values tested are 1, 2, and 5, representing study durations of 1, 2, and 5 years, respectively.
- `interval`: The scheduled interval of study visits. Values tested are 0.25 and 0.5, representing study visits every 3 months and every 6 months, respectively.
- `case`: Specifies the manifold used to generate the data. 
- `obs_noise`: The standard deviation of the independent Gaussian observation-level noise. Values tested include {0.05, 0.1, 0.25, 0.5}.
- `amplitude_noise`: The standard deviation of Gaussian noise applied to the amplitude of the manifold at each time point. Values tested include {0.05, 0.1, 0.25, 0.5}.
- `period_noise`: The standard deviation of Gaussian noise applied to the period of the manifold at each time point. Values tested include {0.05, 0.1, 0.25, 0.5}.
- `time_trend`: Specifies systematic longitudinal changes in the manifold amplitude. Possible values include "constant", indicating that the manifold does not change over time, "linear", representing a constant rate of change in manifold amplitude, and "quadratic", indicating a quadratic rate of change in the manifold amplitude. To approximate trends observed when the brain atrophies, all changes in amplitude are downward.
- `time_change`: Scales the longitudinal amplitude changes. Values tested include {0, 0.05, 0.1, 0.25, 0.5}.

The sample size at each time point is set to 1,000. A factorial design was used to specify combinations of parameter values, with each simulation case being run a total of 5,760 times.

## Simulation Analysis Pipeline

For our purposes, analysis of the simulated data consists of the following steps:

- Preprocessing: Standardize observations to fall between -1 and 1 and augment with polar or spherical coordinates if needed.
- Manifold learning: Fit the LPME, PME and Principal Curve/Surface algorithms to the observed data.
- Model evaluation: For each model, obtain reconstructed estimates of the original observations. Compute the mean squared distance between these estimates and the D-dimensional observations generated from the true manifold.

Functions used to obtain the algorithm-specific reconstructions are provided in the code chunk below:

Functions specifying each step of the analysis process can be found at `code/functions/simulations/preprocess_data.R`, `code/functions/simulations/fit_models.R`, and `code/functions/simulations/evaluate_models.R`. An additional function used to generate plots to display the resulting model estimates can be found at `code/functions/simulations/display_results.R`.

Demonstrations of this analysis pipeline for several simulation cases can be found below.

### Simulation Demonstrations

```{r, eval=FALSE}
#| label: demo_case1
set.seed(100)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 1,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

sim_processed <- preprocess_data(sim_df, case = 1, d = 1, D = 2)
sim_models <- fit_models(sim_processed, case = 1, d = 1, D = 2)
model_errors <- evaluate_models(
  sim_processed,
  sim_models,
  case = 1,
  d = 1,
  D = 2
)
model_errors
result_plot <- display_results(
  sim_processed,
  sim_models,
  case = 1,
  d = 1,
  D = 2
)
result_plot
```

```{r, eval=FALSE}
#| label: demo_case3
set.seed(300)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 3,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

sim_processed <- preprocess_data(sim_df, case = 3, d = 1, D = 2)
sim_models <- fit_models(sim_processed, case = 3, d = 1, D = 2)
model_errors <- evaluate_models(
  sim_processed,
  sim_models,
  case = 3,
  d = 1,
  D = 2
)
model_errors

result_plot <- display_results(
  sim_processed,
  sim_models,
  case = 3,
  d = 1,
  D = 2
)
result_plot
```

```{r, eval=FALSE}
#| label: demo_case7
set.seed(700)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 7,
  obs_noise = 0.1,
  amplitude_noise = 0.01,
  period_noise = 0.01,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

sim_processed <- preprocess_data(sim_df, case = 7, d = 2, D = 3)
sim_models <- fit_models(sim_processed, case = 7, d = 2, D = 3)
model_errors <- evaluate_models(
  sim_processed,
  sim_models,
  case = 7,
  d = 2,
  D = 3
)
model_errors

result_plot <- display_results(
  sim_processed,
  sim_models,
  case = 7,
  d = 2,
  D = 3
)
result_plot
```

```{r, eval=FALSE}
#| label: demo_case8
set.seed(800)
sim_data <- simulate_data(
  duration = 1,
  interval = 0.25,
  case = 8,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

sim_processed <- preprocess_data(sim_df, case = 8, d = 2, D = 3)
sim_models <- fit_models(sim_processed, case = 8, d = 2, D = 3)
model_errors <- evaluate_models(
  sim_processed,
  sim_models,
  case = 8,
  d = 2,
  D = 3
)
model_errors

result_plot <- display_results(
  sim_processed,
  sim_models,
  case = 8,
  d = 2,
  D = 3
)
result_plot
```

## Running Simulations

The code chunks below specify the combinations of simulation parameters and runs the simulations in parallel. Note that due to computational demands, these simulations are not run in the process of compiling this notebook. This was handled by the `code/run_simulations.R` script. However, the code chunk below will yield identical results, and those same results are discussed later in the notebook.

```{r, eval=FALSE}
#| label: simulation_parameters

# set possible values of simulation parameters as discussed above
durations <- c(1, 2, 5)
intervals <- c(0.25, 0.5)
cases <- 1:8
obs_noises <- c(0.05, 0.1, 0.25, 0.5)
amplitude_noises <- c(0.05, 0.1, 0.25, 0.5)
period_noises <- c(0.05, 0.1, 0.25, 0.5)
time_trends <- c("constant", "linear", "quadratic")
time_changes <- c(0, 0.05, 0.1, 0.25, 0.5)

# create dataframe of all combinations of parameter values
parameter_df <- expand_grid(
  duration = durations,
  interval = intervals,
  case = cases,
  obs_noise = obs_noises,
  amplitude_noise = amplitude_noises,
  period_noise = period_noises,
  time_trend = time_trends,
  time_change = time_changes
)

# create empty dataframe to store error estimates
error_df <- tibble(
  lpme_error = vector(mode = "numeric", length = nrow(parameter_df)),
  pme_error = vector(mode = "numeric", length = nrow(parameter_df)),
  pc_error = vector(mode = "numeric", length = nrow(parameter_df)),
  data_error = vector(mode = "numeric", length = nrow(parameter_df))
)

# save the seed state that was set when each simulation was run
seed_states <- list()
```

```{r, eval=FALSE}
#| label: run_simulations
print("Starting Simulations...")

cl <- makeCluster(parallel::detectCores() - 1)
# unclear why, but registerDoSNOW worked better when running with slurm
registerDoSNOW(cl)

# the registerDoRNG function allows reproducibility
# of the simulations when run in parallel
registerDoRNG(611492)

pb  <- txtProgressBar(max = nrow(parameter_df), style = 3)
progress <- function(n) setTxtProgressBar(pb, n)
opts <- list(progress = progress)
error_df <- foreach(
  row_idx = seq_len(nrow(parameter_df)),
  .export = c(
    "simulate_data",
    "preprocess_data",
    "fit_models",
    "evaluate_models",
    "display_results"
  ),
  .packages = c("tidyverse", "pme"),
  .combine = rbind,
  .options.snow = opts
) %dopar% {
  seed_val <- .Random.seed
  seed_states[[row_idx]] <- seed_val

  duration_value <- parameter_df$duration[row_idx]
  interval_value <- parameter_df$interval[row_idx]
  case_value <- parameter_df$case[row_idx]
  obs_noise_value <- parameter_df$obs_noise[row_idx]
  amplitude_noise_value <- parameter_df$amplitude_noise[row_idx]
  period_noise_value <- parameter_df$period_noise[row_idx]
  trend_value <- parameter_df$time_trend[row_idx]
  change_value <- parameter_df$time_change[row_idx]

  # additional stability is needed for Swiss roll and spherical cases
  # to maintain structural integrity
  if (case_value == 7) {
    amplitude_noise_value <- amplitude_noise_value / 10
    period_noise_value <- period_noise_value / 10
  } else if (case_value == 8) {
    period_noise_value <- period_noise_value / 10
  }
